Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-06-20 article-models-ttdp.tex
The first version of the set of rules for the dependent function constructor is finished

\subsubsection{Dependent products are compatible with extension}
\begin{align}
& \inference
  { \jfam{{{{\Gamma}{A}}{B}}{P}}{Q}
    }
  { \jfameq
      {\Gamma}
      {\sprd{{A}{B}}{{P}{Q}}}
      {\ctxext{\sprd{A}{\sprd{B}{P}}}{\sprd{A}{\sprd{B}{Q}}}}
    }
  \\
& \inference
  { \jterm{{{\Gamma}{A}}{B}}{P}{f}
    }
  { \jtermeq
      {\Gamma}
      {\sprd{A}{\sprd{B}{P}}}
      {\slam{{A}{B}}{P}{f}}
      {\slam{A}{\sprd{B}{P}}{\slam{B}{P}{f}}}
    }
\end{align}

@@

\subsubsection{Dependent products are compatible with the identity functions}

>>>>

\subsubsection{Dependent products are compatible with extension}
The following rules describe what happens when a dependent function constructor
is taken over an extension

\begin{align*}
& \inference
  { \jfam{{\Gamma}{A}}{P}
    \jfam{{{\Gamma}{A}}{P}}{Q}
    }
  { \jfameq
      {\Gamma}
      {\mprd{\ctxext{A}{P}}{Q}}
      {\mprd{A}{\mprd{P}{Q}}}
    }
  \\
& \inference
  { \jfam{{\Gamma}{A}}{P}
    \jfam{{{{\Gamma}{A}}{P}}{Q}}{R}
    }
  { \jfameq
      {{\Gamma}{\mprd{\ctxext{A}{P}}{Q}}}
      {\mprd{\ctxext{A}{P}}{R}}
      {\mprd{A}{\mprd{P}{R}}}
    }
  \\
& \inference
  { \jfam{{\Gamma}{A}}{P}
    \jterm{{{{\Gamma}{A}}{P}}{Q}}{R}{h}
    }
  { \jtermeq
      {{\Gamma}{\mprd{\ctxext{A}{P}}{Q}}}
      {\mprd{\ctxext{A}{P}}{R}}
      {\slam{\ctxext{A}{P}}{R}{h}}
      {\slam{A}{\mprd{P}{R}}{\slam{P}{R}{h}}}
    }
  \\
& \inference
  { \jfam{{\Gamma}{A}}{P}
    \jterm{{\Gamma}{\mprd{\ctxext{A}{P}}{Q}}}{\mprd{\ctxext{A}{P}}{R}}{h}
    }
  { \jtermeq
      {{{\Gamma}{{A}{P}}}{Q}}
      {R}
      {\sev{\ctxext{A}{P}}{R}{h}}
      {\sev{P}{R}{\sev{A}{\mprd{P}{R}}{h}}}
    }
\end{align*}

When the dependent product constructor is applied to an extension, we get the
following:

\begin{align*}
& \inference
  { \jfam{{{\Gamma}{A}}{P}}{Q}
    }
  { \jfameq
      {\Gamma}
      {\mprd{A}{\ctxext{P}{Q}}}
      {\ctxext{\mprd{A}{P}}{\mprd{A}{Q}}}
    }
  \\
& \inference
  { \jfam{{{{\Gamma}{A}}{P}}{Q}}{R}
    }
  { \jfameq
      {{\Gamma}{\mprd{A}{P}}}
      {\mprd{A}{\ctxext{Q}{R}}}
      {\ctxext{\mprd{A}{Q}}{\mprd{A}{R}}}
    }
\end{align*}
We show in \autoref{lem:slam-ext} that the remaining two properties
involving lambda abstraction and evaluation are derivable.

@@

\subsubsection{The dependent function constructor is compatible with weakening}
\begin{align*}
& \inference
  { \jfam{{{\Gamma}{A}}{P}}{Q}
    \jfam{{{\Gamma}{Q}}{P}}{R}
    }
  { \jfameq
      {{{\Gamma}{\mprd{A}{P}}}{\mprd{A}{Q}}}
      {\mprd{A}{\ctxwk{Q}{R}}}
      {\ctxwk{\mprd{A}{Q}}{\mprd{A}{R}}}
    }
  \\
& \inference
  { \jfam{{{\Gamma}{A}}{P}}{Q}
    \jfam{{{{\Gamma}{P}}{Q}}{R}}{S}
    }
  { \jfameq
      {{{{\Gamma}{\mprd{A}{P}}}{\mprd{A}{Q}}}{\mprd{A}{\ctxwk{Q}{R}}}}
      {\mprd{A}{\ctxwk{Q}{S}}}
      {\ctxwk{\mprd{A}{Q}}{\mprd{A}{S}}}
    }
  \\
& \inference
  { \jfam{{{\Gamma}{A}}{P}}{Q}
    \jterm{{{{\Gamma}{P}}{Q}}{R}}{S}{k}
    }
  { \jtermeq
      {{{{\Gamma}{\mprd{A}{P}}}{\mprd{A}{Q}}}{\mprd{A}{\ctxwk{Q}{R}}}}
      {\mprd{A}{\ctxwk{Q}{S}}}
      {\slam{A}{\ctxwk{Q}{S}}{\ctxwk{Q}{k}}}
      {\ctxwk{\mprd{A}{Q}}{\slam{A}{S}{k}}}
    }
  \\
& \inference
  { \jfam{{{\Gamma}{A}}{P}}{Q}
    \jfam{{{{\Gamma}{P}}{Q}}{R}}{S}
    \jterm{{{\Gamma}{\mprd{A}{P}}}{\mprd{A}{R}}}{\mprd{A}{S}}{k}
    }
  { \jtermeq
      {{{{{\Gamma}{A}}{P}}{Q}}{\ctxwk{Q}{R}}}
      {\ctxwk{Q}{S}}
      {\sev{A}{\ctxwk{Q}{S}}{\ctxwk{\mprd{A}{Q}}{k}}}
      {\ctxwk{Q}{\sev{A}{S}{k}}}
    }
\end{align*}

\subsubsection{The dependent function constructor is compatible with substitution}
\begin{align*}
& \inference
  { \jterm{{{\Gamma}{A}}{P}}{Q}{g}
    \jfam{{{{\Gamma}{A}}{P}}{Q}}{R}
    }
  { \jfameq
      {{\Gamma}{\mprd{A}{P}}}
      {\mprd{A}{\subst{g}{R}}}
      {\subst{\slam{A}{Q}{g}}{\mprd{A}{R}}}
    }
  \\
& \inference
  { \jterm{{{\Gamma}{A}}{P}}{Q}{g}
    \jfam{{{{{\Gamma}{A}}{P}}{Q}}{R}}{S}
    }
  { \jfameq
      {{{\Gamma}{\mprd{A}{P}}}{\mprd{A}{\subst{g}{R}}}}
      {\mprd{A}{\subst{g}{S}}}
      {\subst{\slam{A}{Q}{g}}{\mprd{A}{S}}}
    }
  \\
& \inference
  { \jterm{{{\Gamma}{A}}{P}}{Q}{g}
    \jterm{{{{{\Gamma}{A}}{P}}{Q}}{R}}{S}{k}
    }
  { \jtermeq
      {{{\Gamma}{\mprd{A}{P}}}{\mprd{A}{\subst{g}{R}}}}
      {\mprd{A}{\subst{g}{S}}}
      {\slam{A}{\subst{g}{S}}{\subst{g}{k}}}
      {\subst{\slam{A}{Q}{g}}{\slam{A}{S}{k}}}
    }
  \\
& \inference
  { \jterm{{{\Gamma}{A}}{P}}{Q}{g}
    \jterm{{\Gamma}{\mprd{A}{\ctxext{{P}{Q}}{R}}}}{\mprd{A}{S}}{k}
    }
  { \jtermeq
      {{{{\Gamma}{A}}{P}}{\subst{g}{R}}}
      {\subst{g}{S}}
      {\sev{A}{\subst{g}{S}}{\subst{\slam{A}{Q}{g}}{k}}}
      {\subst{g}{\sev{A}{S}{k}}}
    }
\end{align*}

\subsubsection{The dependent function constructor is compatible with the identity terms}
\begin{align*}
& \inference
  { \jfam{{{\Gamma}{A}}{P}}{Q}
    }
  { \jtermeq
      {{{\Gamma}{\mprd{A}{P}}}{\mprd{A}{Q}}}
      {\ctxwk{\mprd{A}{Q}}{\mprd{A}{Q}}}
      {\slam{A}{\ctxwk{Q}{Q}}{\idtm{Q}}}
      {\idtm{\mprd{A}{Q}}}
    }
  \\
& \inference
  { \jfam{{{\Gamma}{A}}{P}}{Q}
    }
  { \jtermeq
      {{{{\Gamma}{A}}{P}}{Q}}
      {\ctxwk{Q}{Q}}
      {\sev{A}{\ctxwk{Q}{Q}}{\idtm{\mprd{A}{Q}}}}
      {\idtm{Q}}
    }
\end{align*}

\subsection{Consequences of the rules for the dependent function constructors}

\begin{lem}\label{lem:slam-ext}
The inference rules asserting that lambda abstraction is compatible with term
extension
\begin{align*}
& \inference
  { \jterm{{{\Gamma}{A}}{P}}{{Q}{R}}{h}
    }
  { \jtermeq
      {{\Gamma}{\mprd{A}{P}}}
      {\mprd{A}{\ctxext{Q}{R}}}
      {\cprojfst{\mprd{A}{Q}}{\mprd{A}{R}}{\slam{A}{\ctxext{Q}{R}}{h}}}
      {\slam{A}{Q}{\cprojfst{Q}{R}{h}}}
    }
  \\
& \inference
  { \jterm{{{\Gamma}{A}}{P}}{{Q}{R}}{h}
    }
  { \jtermeq
      {{\Gamma}{\mprd{A}{P}}}
      {\mprd{A}{\ctxext{Q}{R}}}
      {\cprojfst{\mprd{A}{Q}}{\mprd{A}{R}}{\slam{A}{\ctxext{Q}{R}}{h}}}
      {\slam{A}{R}{\cprojsnd{Q}{R}{h}}}
    }
\intertext{and the inference rules asserting that evaluation is compatible with
term extension}
& \inference
  { \jterm{{\Gamma}{\mprd{A}{P}}}{\mprd{A}{\ctxext{Q}{R}}}{h}
    }
  { \jtermeq
      {{{\Gamma}{A}}{P}}
      {{Q}{R}}
      {\cprojfst{Q}{R}{\sev{A}{\ctxext{Q}{R}}{h}}}
      {\sev{A}{Q}{\cprojfst{\mprd{A}{Q}}{\mprd{A}{R}}{h}}}
    }
  \\
& \inference
  { \jterm{{\Gamma}{\mprd{A}{P}}}{\mprd{A}{\ctxext{Q}{R}}}{h}
    }
  { \jtermeq
      {{{\Gamma}{A}}{P}}
      {{Q}{R}}
      {\cprojsnd{Q}{R}{\sev{A}{\ctxext{Q}{R}}{h}}}
      {\sev{A}{R}{\cprojsnd{\mprd{A}{Q}}{\mprd{A}{R}}{h}}}
    }
\end{align*}
are valid.
\end{lem}

\begin{proof}
We have the judgmental equalities
\begin{align*}
\cprojfst{\mprd{A}{Q}}{\mprd{A}{R}}{\slam{A}{\ctxext{Q}{R}}{h}}
& \jdeq
  \subst{\slam{A}{\ctxext{Q}{R}}{h}}{\ctxwk{\mprd{A}{R}}{\idtm{\mprd{A}{Q}}}}
  \\
& \jdeq 
  \subst{\slam{A}{\ctxext{Q}{R}}{h}}{\ctxwk{\mprd{A}{R}}{\slam{A}{\ctxwk{Q}{Q}}{\idtm{Q}}}}
  \\
& \jdeq
  \subst{\slam{A}{\ctxext{Q}{R}}{h}}{\slam{A}{\ctxwk{R}{{Q}{Q}}}{\ctxwk{R}{\idtm{Q}}}}
  \\
& \jdeq
  \subst{\slam{A}{\ctxext{Q}{R}}{h}}{\slam{A}{\ctxwk{\ctxext{Q}{R}}{Q}}{\ctxwk{R}{\idtm{Q}}}}
  \\
& \jdeq
  \slam{A}{Q}{\subst{h}{\ctxwk{R}{\idtm{Q}}}}
  \\
& \jdeq
  \slam{A}{Q}{\cprojfst{Q}{R}{h}}
\end{align*}
\end{proof}

\begin{rmk}
It follows from these rules that on the projections (i.e.~on the fibrations),
lambda abstraction and evaluation are each other's inverses even though we
have not even postulated that evaluation is a retraction of lambda 
abstraction. Thus, in combination with univalence this should give that
they are homotopically inverse to each other.
\end{rmk}
